WHOOMP! (There It Is) \\ Rapid Bayesian position reconstruction for gravitational\nobreakdashes-wave transients

### 2014-09-17 ms.tex
Remove unnecessary phrase
Rewrite intro

Ground-based \ac{GW} interferometers are entering the advanced detector era, positioning themselves to make profound discoveries about the universe. The \ac{aLIGO} and Vigo detectors will be part of a worldwide network that also includes KAGRA and hopefully LIGO-India. Coalescing neutron star binaries are among the most likely sources, with 0.4\nobreakdashes--400 detections per year~\cite{rates}. In addition to being an efficient source of \acp{GW}, the \ac{NS} may be tidally shredded before merging, providing fuel for an electromagnetic counterpart. The short-lived accretion flow may power a collimated relativistic jet, resulting in a short \ac{GRB} and an X-ray afterglow if it is aligned with the line of sight. An optical afterglow may follow minutes to hours later as the jet plows into the interstellar medium. As surrounding neutron-rich ejecta decay radioactively, an optical `kilonova' may be visible after $\sim$1 day \cite{metzger:2010}. Finally, as the ejecta plow into the interstellar medium, a faint radio afterglow could occur on a timescale of weeks to years \cite{Nakar:2011cw}.

There is therefore a strong case for performing electromagnetic followups of these sources of \acp{GW}. Indeed, the final science run of the intitial LIGO and Virgo instruments saw the first joint search for \acp{GW} from compact binaries and their electromagnetic counterparts \cite{abadie2012first}.  As we prepare for the next generation of these searches there is a need for determining the location of the source as precisely as possible in as little time as possible.

In this paper we present a rapid and accurate method of sky localization that makes use of Bayesian methods. It differs from existing techniques in several important ways.  In the first place, we fix the intrinsic parameters to their \ac{ML} estimates, as provided by the detection pipeline.  This reduces the dimensionality of the parameter space we need to sample. In addition, the technique takes as its input the \ac{ML} estimates of the extrinsic parameters, instead of the $h(t)$ time series. This makes the likelihood itself much faster to evaluate. Finally, instead of of using \ac{MCMC} or some similar method for statistical sampling, we make use of a deterministic quadrature scheme. We call this technique \ac{BAYESTAR}%
%
\footnote{A pun on the Cylon battleships in the American television series Battlestar Galactica. The defining characteristic of the Cylons is that they repeatedly defeat humanity by using their superhuman information\nobreakdashes-gathering ability to coordinate overwhelming forces. The name also suggests that, like the Cylons, \ac{GW} detectors may some day rise against us humans.}%
%
\footnote{We do not like to mention the final `L' in the acronym, because then it would be pronounced BAYESTARL, which sounds stupid.}%
. It is unique in that it bridges the detection and parameter estimation of \ac{GW} signals, two tasks that have until now involved very different numerical methods and time scales. We expect that \ac{BAYESTAR} will take on a key role in observing \ac{CBC} events in both \ac{GW} and optical channels during the Advanced \ac{LIGO} era.

@@

A \ac{CBC} source is specified by a vector of extrinsic parameters describing its position and orientation and intrinsic parameters describing the physical properties of the binary components:
%
\begin{equation}
    \bm\theta = \left[
    \begin{array}{cl@{\quad}p{4cm}lp{3.75cm}}
        \alpha & \rdelim]{11}{0mm} & right ascension & \rdelim\}{7}{1mm} & \multirow{7}{2cm}{extrinsic parameters, $\bm\theta_\mathrm{ex}$} \\
        \delta && declination & \\
        r && distance & \\
        t_\oplus && arrival time at geocenter & \\
        \iota && inclination angle & \\
        \psi && polarization angle & \\
        \phi_c && coalescence phase & \\
        \cline{1-1}\cline{3-3}
        m_1 && first component's mass & \rdelim\}{4}{1mm} & \multirow{4}{2cm}{intrinsic parameters, $\bm\theta_\mathrm{in}$.}\\
        m_2 && second component's mass & \\
        \mathbf S_1 && first component's spin & \\
        \mathbf S_2 && second component's spin & \\
    \end{array}\right.
\end{equation}
%
This list of parameters involves some simplifying assumptions. Eccentricity is omitted: although it does play a major role in the evolution and waveforms of \ac{NSBH} and \ac{BBH} sources formed by dynamical capture~\cite{PhysRevD.87.043004}, \ac{BNS} systems formed by binary stellar evolution should circularize due to tidal interaction~\cite{0004-637X-572-1-407} and later \ac{GW} emission~\cite{PhysRev.136.B1224} long before the inspiral enters \ac{LIGO}'s frequency range of $\sim$10\nobreakdashes--1000~kHz. Tidal deformabilities of the \acp{NS} are omitted because the signal imprinted by the companions' material properties is so small that it will only be detectable by an Einstein Telescope\nobreakdashes-class \ac{GW} observatory~\cite{PhysRevD.81.123016}. Furthermore, in \ac{GW} detection efforts, especially those focused on \ac{BNS} systems, the component spins $\mathbf{S}_1$ and $\mathbf{S}_2$ are often assumed to be nonprecessing and aligned with the system's total angular momentum and condensed to a single scalar parameter $\chi$, or even neglected entirely, $\mathbf{S}_1 = \mathbf{S}_2 = 0$.

@@

The choice of prior is open to one's astrophysical preconceptions, but during \ac{S6} when \ac{LIGO}'s Bayesian \ac{CBC} parameter estimation pipelines were pioneered, the prior was taken to be isotropic in sky location and binary orientation, and uniform in volume, arrival time, and the component masses~\cite{2013arXiv1304.1775T}.

@@

Bayesian parameter estimation has many advantages, including broad generality and the ability to make probabilistically meaningful statements even with very low \ac{SNR} measurements. However, in problems of even modest complexity, the marginalization step involves many-dimensional, ill-behaved integrals. The powerful \ac{MCMC} integration technique has become almost synonymous with Bayesian inference. Though powerful, \ac{MCMC} is inherently non-deterministic and resistant to parallelization, as well as (at least historically) slow. With the most sophisticated \ac{CBC} parameter estimation codes, it still takes days to process a single event. This delay is undesirable for planning targeted \ac{EM} follow\nobreakdashes-up searches of \ac{LIGO} events.

@@

\begin{acknowledgements}
\marginpar{This URL is not specific enough.}Source code for \ac{BAYESTAR} is available on the \ac{LIGO} \acl{DASWG} web site at \url{http://www.lsc-group.phys.uwm.edu/daswg/projects/lalsuite.html}.

@@

\ac{LIGO} was constructed by the California Institute of Technology and Massachusetts Institute of Technology with funding from the \ac{NSF} and operates under cooperative agreement PHY\nobreakdashes-0107417. Some results were produced on the NEMO computing cluster operated by the Center for Gravitation and Cosmology at University of Wisconsin\nobreakdashes--Milwaukee under \ac{NSF} Grants PHY\nobreakdashes-0923409 and PHY\nobreakdashes-0600953. This research is supported by the \ac{NSF} through a Graduate Research Fellowship to L.S. This paper has \ac{LIGO} Document Number \ac{LIGO}\nobreakdashes-PXXXXXXX\nobreakdashes-vX.
\end{acknowledgements}

>>>>

The Advanced LIGO and Virgo ground-based \ac{GW} detectors will soon begin taking data \cite{LIGOObservingScenarios} in their `Advanced' configuration with significantly greater sensitivity than their initial runs through 2006. They will form the first parts of a sensitive global network to be later augmented by the Japanese KAGRA facility \cite{LCGT,KAGRAInterferometerDesign} and LIGO\nobreakdashes--India \cite{LIGOIndia}. Their most likely sources of \acp{GW} include 0.4\nobreakdashes--400 detections binary neutron star mergers per year~\cite{rates}.

These systems are both efficient sources of \acp{GW} and potential sources of detectable electromagnetic transients \citep{MostPromisingEMCounterpart}. If one of the binary companions is a \ac{NS}, then it may be tidally shredded before merging. If the system's orbital axis is aligned with the line of sight, this may produce a short, hard \acp{GRB}
\cite{1986ApJ...308L..43P,1989Natur.340..126E,1992ApJ...395L..83N,2011ApJ...732L...6R}. An X\nobreakdashes-ray and optical afterglow may follow minutes to hours later as the jet plows into the interstellar medium. As surrounding neutron-rich ejecta decay radioactively, an optical `kilonova' may be visible after $\sim$1 day \cite{metzger:2010}. Finally, as the ejecta itself plow into the interstellar medium, a faint radio afterglow could rise and fall on timescales of weeks to years \cite{Nakar:2011cw}.

The on\nobreakdashes-axis optical afterglows and the recently hypothesized kilonova precursors \cite{KilonovaPrecursor} may be among the most straightforward electromagnetic components to pinpoint with existing and planned optical transient facilities such as the Zwicky Transient Facility~\citep{ZTF} and BlackGEM\footnote{\url{https://www.astro.ru.nl/wiki/research/blackgemarray}}. However, both of these signatures fade rapidly, on a timescale of hours. Because \ac{GW} detectors have inherently coarse sky position reconstruction ability \cite{FairhurstTriangulation,WenLocalizationAdvancedLIGO,FairhurstLocalizationAdvancedLIGO,2011PhRvD..84j4020V,RodriguezBasicParameterEstimation,NissankeLocalization,NissankeKasliwalEMCounterparts,KasliwalTwoDetectors,Grover:2013,SiderySkyLocalizationComparison,FirstTwoYears}, detecting these transients may depend upon rapidly calculating and disseminating \ac{GW} probability sky maps.

The final science run of the initial LIGO and Virgo instruments saw the first joint search for \ac{GW} and electromagnetic emission from compact binaries. This involved several advances in the \ac{GW} data analysis, including the first real\nobreakdashes-time detection and rapid sky localization pipeline \cite{abadie2012first} and the first practical \ac{MCMC} parameter estimation analysis of \ac{GW} signals \cite{S6PE}. Despite these technical achievements, there was an undesirable tradeoff between the speed of the rapid localization and the full parameter estimation: the former could analyze a detection candidate in minutes, whereas the latter took half a day; the latter decreased the area on the sky by a factor of 1/20 over the former but took 1000 times as long to run \cite{SiderySkyLocalizationComparison}.

For the purpose of electromagnetic observations, we would like to have a nearly final sky localization within a timescale of minutes to hours. To that end, we have devised a rapid and accurate Bayesian sky localization method that takes mere seconds but achieves approximately the same accuracy as the full \ac{MCMC} analysis. We call this algorithm \ac{BAYESTAR}%
%
\footnote{A pun on the Cylon battleships in the American television series Battlestar Galactica. The defining characteristic of the Cylons is that they repeatedly defeat humanity by using their superhuman information\nobreakdashes-gathering ability to coordinate overwhelming forces. The name also suggests that, like the Cylons, \ac{GW} detectors may some day rise against us humans.}%
%
\footnote{We do not like to mention the final `L' in the acronym, because then it would be pronounced BAYESTARL, which sounds stupid.}%
. It differs from existing techniques in several important ways. In the first place, we treat the matched-filter detection pipeline as a measurement system in and of itself, treating the point\nobreakdashes-parameter estimates that it provides as the experimental input data rather than the full \ac{GW} time series that is used by the \ac{MCMC} analysis. This drastically reduces the dimensionality of both the data and the signal model. It also permits us to avoid directly computing the expensive post-Newtonian model waveforms, making the likelihood itself much faster to evaluate. Finally, instead of of using \ac{MCMC} or some similar method for statistical sampling, we make use of a deterministic quadrature scheme. This algorithm is unique in that it bridges the detection and parameter estimation of \ac{GW} signals, two tasks that have until now involved very different numerical methods and time scales. We expect that \ac{BAYESTAR} will take on a key role in observing \ac{CBC} events in both \ac{GW} and optical channels during the Advanced LIGO era.

@@

A \ac{CBC} source is specified by a vector of extrinsic parameters describing its position and orientation and intrinsic parameters describing the physical properties of the binary components:
%
\begin{equation}
    \bm\theta = \left[
    \begin{array}{cl@{\quad}p{4cm}lp{3.75cm}}
        \alpha & \rdelim]{11}{0mm} & right ascension & \rdelim\}{7}{1mm} & \multirow{7}{2cm}{extrinsic parameters, $\bm\theta_\mathrm{ex}$} \\
        \delta && declination & \\
        r && distance & \\
        t_\oplus && arrival time at geocenter & \\
        \iota && inclination angle & \\
        \psi && polarization angle & \\
        \phi_c && coalescence phase & \\
        \cline{1-1}\cline{3-3}
        m_1 && first component's mass & \rdelim\}{4}{1mm} & \multirow{4}{2cm}{intrinsic parameters, $\bm\theta_\mathrm{in}$.}\\
        m_2 && second component's mass & \\
        \mathbf S_1 && first component's spin & \\
        \mathbf S_2 && second component's spin & \\
    \end{array}\right.
\end{equation}
%
This list of parameters involves some simplifying assumptions. Eccentricity is omitted: although it does play a major role in the evolution and waveforms of \ac{NSBH} and \ac{BBH} sources formed by dynamical capture~\cite{PhysRevD.87.043004}, \ac{BNS} systems formed by binary stellar evolution should circularize due to tidal interaction~\cite{0004-637X-572-1-407} and later \ac{GW} emission~\cite{PhysRev.136.B1224} long before the inspiral enters LIGO's frequency range of $\sim$10\nobreakdashes--1000~kHz. Tidal deformabilities of the \acp{NS} are omitted because the signal imprinted by the companions' material properties is so small that it will only be detectable by an Einstein Telescope\nobreakdashes-class \ac{GW} observatory~\cite{PhysRevD.81.123016}. Furthermore, in \ac{GW} detection efforts, especially those focused on \ac{BNS} systems, the component spins $\mathbf{S}_1$ and $\mathbf{S}_2$ are often assumed to be nonprecessing and aligned with the system's total angular momentum and condensed to a single scalar parameter $\chi$, or even neglected entirely, $\mathbf{S}_1 = \mathbf{S}_2 = 0$.

@@

The choice of prior is open to one's astrophysical preconceptions, but during \ac{S6} when LIGO's Bayesian \ac{CBC} parameter estimation pipelines were pioneered, the prior was taken to be isotropic in sky location and binary orientation, and uniform in volume, arrival time, and the component masses~\cite{2013arXiv1304.1775T}.

@@

Bayesian parameter estimation has many advantages, including broad generality and the ability to make probabilistically meaningful statements even with very low \ac{SNR} measurements. However, in problems of even modest complexity, the marginalization step involves many-dimensional, ill-behaved integrals. The powerful \ac{MCMC} integration technique has become almost synonymous with Bayesian inference. Though powerful, \ac{MCMC} is inherently non-deterministic and resistant to parallelization, as well as (at least historically) slow. With the most sophisticated \ac{CBC} parameter estimation codes, it still takes days to process a single event. This delay is undesirable for planning targeted \ac{EM} follow\nobreakdashes-up searches of LIGO events.

@@

\begin{acknowledgements}
\marginpar{This URL is not specific enough.}Source code for \ac{BAYESTAR} is available on the LIGO \acl{DASWG} web site at \url{http://www.lsc-group.phys.uwm.edu/daswg/projects/lalsuite.html}.

@@

LIGO was constructed by the California Institute of Technology and Massachusetts Institute of Technology with funding from the \ac{NSF} and operates under cooperative agreement PHY\nobreakdashes-0107417. Some results were produced on the NEMO computing cluster operated by the Center for Gravitation and Cosmology at University of Wisconsin\nobreakdashes--Milwaukee under \ac{NSF} Grants PHY\nobreakdashes-0923409 and PHY\nobreakdashes-0600953. This research is supported by the \ac{NSF} through a Graduate Research Fellowship to L.S. This paper has LIGO Document Number LIGO\nobreakdashes-PXXXXXXX\nobreakdashes-vX.
\end{acknowledgements}
